---
title: "RMarkdown"
output:
  html_document: default
  word_document: default
  pdf_document: default
date: '2022-07-01'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

An RMarkdown file will allow us to weave markdown text with chunks of R code to be evaluated and output content like tables and plots. The RMarkdown file can be output to different format such as HTML, PDF and Word. Further, we can use LaTeX codes in the RMarkdown file to prepare a professional document.  
<br />
First we demonstrate how to create a simple RMarkdown file with output to HTML format. Next, we will demonstrate how to use LaTeX code in RMarkdown file and output to PDF format.  



### Procedures to create a RMarkdown file  

1. File -> New File -> RMarkdown ... -> Document of output format HTML, click `OK`.
2. A template RMarkdown file is created. There are white and grey sections. R code is in grey sections, and other text is in white.
3. "Knit HTML" by clicking the blue yarn at the top of the RMarkdown file and the HTML document will be created.


### Codes used in the RMarkdown file
*   **R code chunks** are surrounded by 3 backticks and `{r LABEL}`. These code chunks are shown in the document and result of the code chunks will be followed.
*   The R code chunks with the option `echo=FALSE`, such as `{r pressure, echo=FALSE}, the printing of the code chunks will be suppressed and only the result will be shown.
*   The code chunks could be printed without evaluation by enclosing the code beginning with 3 backticks and the type of programme, like ````r`, and ended with 3 backticks.
*   The hashtag (#) cause the following text to be displayed larger and in bold.  

<br />
For RStudio built in reference: Go to Help > Markdown Quick Reference.  
<br />
You can get further reference at http://rmarkdown.rstudio.com/

### Knite to PDF and use LaTeX commands
In order to output document in PDF format, we need to install MikTeX and run a R code. These steps just need to perform once.

1.  Install MikTeX.
2.  Run the following R code:
    ```{r eval=FALSE}
       tinytex::install_tinytex()
    ```



    
 


