---
title: "Setup Git & GitHub"
output:
  pdf_document: default
  word_document: default
  html_document: default
date: '2022-06-28'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Interface with GitHub using RStudio

### Procedures
The following steps are needed to setup interface with GitHub using RStudio. The first step is needed to be set up once. The second step is needed when a new repository is created. (Noted: In order to create a nested list, two tabs or 4 spaces are needed to indent for every level. In order to add a blank line, add two spaces at the end of the previous paragraph and followed by the `<br />` command.)

1. One time setup
    i.    Install Git on Windows
    ii.   Create Github account at http://github.com
    iii.  Use the `usethis` package to configure git with global commands.
    
  
    ```r
    install.packages("usethis")
    library(usethis)
    use_git_config(user.name = "stxxxxxxxun", user.email = "stxxxxxxxun@gmail.com")
    ```

    iv.   In RStudio, navigate to: Tools > Global Options > Git/SVN and check whether the filepath is found in "**Git executable**".
    v.    Create a `GitHub` folder in the local machine to keep all the GitHub repositories.  
<br />
2. Create a repository on Github.com
    i.    Go to the account on github.com and click "New repository".
    ii.   Give a simple name and add a description
    iii.  Public or private (private available with paid or education)
    iv.   Select "Initialize this repository with a README.
    v.    Ignore other selection and click "Create repository".  
<br />
3. Clone your repository using RStudio
    i.    Copy the web address of the repository you want to clone.
    ii.   From RStudio, go to New Project (in the File menu).
    iii.  Select Version Control
    iv.   Select Git
    v.    Paste it in the Repository URL field, and click `tab` to autofill the Project Directory name. Make sure to keep the Project Directory Name **THE SAME** as the repository name from the URL.
    vi.   Click `Create Project`.


<br />

Examples in create table

First Header  | Second Header
------------- | -------------
Content Cell  | Content Cell
Content Cell  | Content Cell
Reference Style Links and Images

```{r tables-mtcars}
knitr::kable(iris[1:5, ], caption = 'A caption')
```