---
title: "Linear regression"
output:
  html_document:
    number_sections: true
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Suppose we have three vectors $y$, $x_1$ and $x_2$ and we want to fit the model:

$$Y=\beta_0+\beta_1{x_1}+\beta_2{x_2}+u$$

```{r}
# Preparing the data
x1 <- seq(1,10,length=25)
x2 <- runif(25,3,7)
y <- 4 + 2*x1 + 7*x2 + rnorm(25,0,1)
mydata <- data.frame(y=y, x1=x1, x2=x2)
head(mydata)
```

# Linear regression using the command `lm`

We run the regression using the command `lm`:

```{r}
# Run the regression
out <- lm(y ~ x1 + x2, data = mydata)
out                                   # Basic regression result
names(out)                            # Parameters attached to the object 'out'

# Obtain the particular estimates
out$coefficients                      # All the coefficients
out$coefficients[2]                   # Coefficient for x1
```

More detailed results can be obtained by the commands `summary` and `anova`.

```{r}
# Summary of the regression result
s <- summary(out)            
summary(out)                 # Use just 'summary()' to display the result
# ANOVA table
a <- anova(out)              # Anova table
print(a)                     # Use command 'print' to display the result
```

Different parameters are attached to the object `s` and `a`. We can display the attached parameters by the `names` command. and we can extract different information of by the following commands:\

**Parameters available in the object `summary()`**

```{r}
# Parameters available in the object 'summary()'
names(s)           # Summary of the regression
# Coefficients of the regression
s$coefficients       # Give more details than 'out$coefficients"
s$coefficients[2]    # Coefficient of x1
s$coefficients[2,4]  # Pr(>|t|) for x1
```

**Parameters available in the object `anova()`**
```{r}
# Parameters available in the object 'anova()'
names(a)           # Anova of the regression
a                  # The ANOVA table
a$`Sum Sq`         # Displaying all the sums of squares
a$`Sum Sq`[3]      # Displaying SSR
sum(a$`Sum Sq`)    # Calculating SST
```

Finally, the **AIC statistics** can be obtained by the command `extractAIC`
```{r}
# AIC statistics
extractAIC(out)
```

# Linear regression by the command `lsfit`

Here is the another way to do linear regression.

```{r}
X <- cbind(x1,x2)
temp <- lsfit(X,y)
ls.print(temp)
names(temp)

