---
title: "Basic R operations"
output:
  html_document: default
  word_document: default
  pdf_document: default
date: '2022-07-23'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Key terms

* Library:`foreign`; Command: `read.dta()`
* Library: `haven`; command: `read_sav()`
* Command: `read.table()`
* Library: `pdfetch`
* Library: `quantmod`
* Residual
* Residual sum of squares, _RSS_
* Ordinary least squares, _OLS_

# File and directory management

```{r}
# File and directory Management
# Rscript file: \Rscripts\FileManagement.R

# List all the objects in your local workspace using
ls()

# To demonstrate how to clear an object, we first load an `RData` file into the an R workspace
load(file=url("http://faculty.gvsu.edu/kilburnw/pollingexcerpt.RData"))

# Two objects, `poll` and `poll16` are loaded into the R workspace
ls()

# Clear the objects
rm(poll)
rm(poll16)
ls()

# To clear all the objects in your local workspace, type `rm(list=ls())`
# rm(list=ls())          # Not run


# List all the files in your working directory using
list.files()

# or
dir()

# Get the current working directory
getwd()

# Assign the value of the current working directory to a variable called "old.dir" for later use
old.dir <- getwd()

# Change to a directory by typing the full path
setwd('D:/Documents/GitHub/R4DS/Rscripts')
getwd()

# Return to the parent directory (one level up the present directory)
setwd('..')       # or `setwd('../')`
getwd()

# Change to a subdirectory
setwd('./Rscripts')
getwd()

# Change to a directory by stored variable
setwd(old.dir)
getwd()


# Use dir.create() to create a directory in the current working directory called "testdir".
dir.create('testdir')
dir()

# Create a sub-subdirectory
dir.create('testdir/subtestdir1')

# Change to the subdirectory `testdir1` and create a subdirectory here
setwd('./testdir')
dir()
dir.create('subtestdir2')
dir()

# Moving up one level of the directory (parent directory)
setwd('..')       # or `setwd('../')`
getwd()
dir()

# Remove a directory together will all its subdirectories
unlink("testdir", recursive=TRUE)
dir()
```

# Importing data


### Loading data of various file format

```{r, warning=FALSE}
# Importing data file of various format
# Rscript file: /Rscripts/LoadData.R

# Importing Stata data file into R
# Packages `foreign` required
library(foreign)               # Load the required package

# Input the Stata data file `wage.dta` 
# `wage.dta` is a data file accompany the book "An Introduction to Classical Econometric Theory (2000)" by
# Paul A. Ruud

DTAdata <- read.dta("./Data/wage.dta")
head(DTAdata)

# Further example of importing Stata data file from internet
affairs <- read.dta("http://fmwww.bc.edu/ec-p/data/wooldridge/affairs.dta")
head(affairs)

##########################################################################################################
# Importing SPSS data file into R
# Packages `haven` required
library(haven)               # Load the required package

# Input the SPSS data file `Cats and Dogs.sav` 
# `Cats and Dogs.sav` is a data file accompany the book "Discovering Statistics Using IBM SPSS Statistics
# (2018)" by Andy Field

SPSSdata <- read_sav("./Data/Cats and Dogs.sav")
head(SPSSdata)

##########################################################################################################
# Display the objects created when importing the data files
ls()

# Removing the objects
rm(DTAdata)
rm(SPSSdata)
rm(affairs)
```


### Importing data from internet

The following is adapted from Example 1.13^[See Jim Albert and Maria Rizzo, _R by Example_, Springer, New York, 2012, pp. 30.] (Digits of $\pi$) of the book "R by Example" by Jim Albert and Mraia Rizzo. The data file “PiDigits.dat^[The text file has been downloaded and saved in the sub-directory "Data."]” contains the first 5000 digits of the mathematical constant $\pi =3.1415926535897932384\dots$. The data is one of the Statistical Reference Datasets provided by the National Institute of Standards and Technology (NIST). 2 Documentation is inserted at the top of the file, and the digits start on line 61. We use the `read.table` function with the complete URL (web address) and skip=60 to read the data starting at line 61. The URL should be enclosed in quotes and on a single line; otherwise an error message “cannot open the connection” occurs.

```{r}
#Import data from internet
#Example 1.13 of the book "R by Example"
#Rscript file: \Rscripts\RbyEG_EG1_13.R

# Read in the data
pidigits = read.table(
  "http://www.itl.nist.gov/div898/strd/univ/data/PiDigits.dat",
  skip=60)
head(pidigits)
#Summarize the digits in a table by 'table' function
table(pidigits)
#Calculate the proportions of the digits
prop <- table(pidigits) / 5000
prop
#Standard error (se) of the proportion with known proportion
sqrt(.1*.9 / 5000)
#For unknow proportion, sample se is calculated
se.hat <- sqrt(prop * (1 - prop) / 5000)
se.hat
#The sample proportion plus or minus two standard errors using vectorized arithmetic
round(rbind(prop, se.hat, prop-2*se.hat, prop+2*se.hat), 4)
#A barplots helps to visualize the tabulatd data
barplot(prop, xlab="digit", ylab="proportion")
abline(h = .1)
```

### Getting financial data from Yahoo using the package `pdfetch`

```{r, message=FALSE, warning=FALSE}
# Getting financial data from Yahoo using the package `pdfetch`
# Rscript file: /Rscripts/pdfetch.R

library(pdfetch)       # Load the `pdfetch` package

tickernames <- c("0001.HK","0005.HK")
HKStocks <- pdfetch_YAHOO(tickernames, fields="adjclose",from="2018-01-01")
nrow(HKStocks)
plot(HKStocks$"0001.HK", las=2)
plot(HKStocks$"0005.HK", las=2)
```
  
<br />

### Retrieving economic data from the internet using the package `quantmod`

The package `quantmod` is used to retrieve financial data from the internet such as Yahoo Finance and the Federal Reserve Economic Data (FRED) of Federal Reserve Bak of St. Louis.  
<br />
First, we demonstrate how to retrieve stock prices from Yahoo Finance.

```{r, message=FALSE, warning=FALSE}
# Retrieve economic data from the internet using the `quantmod` package
# Rscript file: \Rscripts\quantmod_Yahoo_US.R

# Retrieve stock price from Yahoo Finance
# Load the required package
library(quantmod)       #Alternatively `require(quantmod)`

#Get Apple stock price
getSymbols("AAPL",from="2008-01-03",to="2015-01-28")

#See the size of the loaded data set
dim(AAPL)

#See the first and last 6 row of data
head(AAPL)
tail(AAPL)

#Obtain time plot of closing price and trading volume
#The subcommand theme="white" set the background of the time plot
#The default is black
chartSeries(AAPL,theme="white")
```

# Find and Count Unique Values

```{r, warning=FALSE, message=FALSE}
library(tidyverse)

nycflights13::flights %>% 
  print(n = 10, width = Inf)

n_distinct(nycflights13::flights$origin)

unique(nycflights13::flights$origin)

table(nycflights13::flights$origin)
```





# Unclassified contents


<br />

**Examples in create table**

First Header  | Second Header
------------- | -------------
Content Cell  | Content Cell
Content Cell  | Content Cell
Reference Style Links and Images


**Example in creating table by `knitr::kable()`**
```{r tables-mtcars}
knitr::kable(iris[1:5, ], caption = 'A caption')
```

**Obtain predicted values, residuals and anova of `lm()`**
```{r , message=FALSE}
# EARNINGS regress on schooling
library(dplyr)
eawe21 <- read.csv('./Data/EAWE21.csv', header=T)   # Import data
lm(EARNINGS ~ S, data = eawe21) %>%
  anova
```
(Please the website [https://rc2e.com/linearregressionandanova][rc2e] for more information on regression output.)

[rc2e]: https://rc2e.com/linearregressionandanova