---
title: "Basic R operations"
output:
  html_document: default
  word_document: default
  pdf_document: default
date: '2022-07-23'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Key terms

* `read.table()`
* `quantmod`
* Disturbance term
* Residual
* Residual sum of squares, _RSS_
* Ordinary least squares, _OLS_


# Importing data

### Importing data from internet

The following is adapted from Example 1.13^[See Jim Albert and Maria Rizzo, _R by Example_, Springer, New York, 2012, pp. 30.] (Digits of $\pi$) of the book "R by Example" by Jim Albert and Mraia Rizzo. The data file “PiDigits.dat^[The text file has been downloaded and saved in the sub-directory "Data."]” contains the first 5000 digits of the mathematical constant $\pi =3.1415926535897932384\dots$. The data is one of the Statistical Reference Datasets provided by the National Institute of Standards and Technology (NIST). 2 Documentation is inserted at the top of the file, and the digits start on line 61. We use the `read.table` function with the complete URL (web address) and skip=60 to read the data starting at line 61. The URL should be enclosed in quotes and on a single line; otherwise an error message “cannot open the connection” occurs.

```{r}
#Import data from internet
#Example 1.13 of the book "R by Example"
#Rscript file: \Rscripts\RbyEG_EG1_13.R

# Read in the data
pidigits = read.table(
  "http://www.itl.nist.gov/div898/strd/univ/data/PiDigits.dat",
  skip=60)
head(pidigits)
#Summarize the digits in a table by 'table' function
table(pidigits)
#Calculate the proportions of the digits
prop <- table(pidigits) / 5000
prop
#Standard error (se) of the proportion with known proportion
sqrt(.1*.9 / 5000)
#For unknow proportion, sample se is calculated
se.hat <- sqrt(prop * (1 - prop) / 5000)
se.hat
#The sample proportion plus or minus two standard errors using vectorized arithmetic
round(rbind(prop, se.hat, prop-2*se.hat, prop+2*se.hat), 4)
#A barplots helps to visualize the tabulatd data
barplot(prop, xlab="digit", ylab="proportion")
abline(h = .1)
```

### Retrieving economic data from the internet using the package `quantmod`

The package `quantmod` is used to retrieve financial data from the internet such as Yahoo Finance and the Federal Reserve Economic Data (FRED) of Federal Reserve Bak of St. Louis.  
<br />
First, we demonstrate how to retrieve stock prices from Yahoo Finance.

```{r, message=FALSE, warning=FALSE}
# Retrieve economic data from the internet using the `quantmod` package
# Rscript file: \Rscripts\quantmod_Yahoo_US.R

# Retrieve stock price from Yahoo Finance
# Load the required package
library(quantmod)       #Alternatively `require(quantmod)`

#Get Apple stock price
getSymbols("AAPL",from="2008-01-03",to="2015-01-28")

#See the size of the loaded data set
dim(AAPL)

#See the first and last 6 row of data
head(AAPL)
tail(AAPL)

#Obtain time plot of closing price and trading volume
#The subcommand theme="white" set the background of the time plot
#The default is black
chartSeries(AAPL,theme="white")
```



<br />

**Examples in create table**

First Header  | Second Header
------------- | -------------
Content Cell  | Content Cell
Content Cell  | Content Cell
Reference Style Links and Images


**Example in creating table by `knitr::kable()`**
```{r tables-mtcars}
knitr::kable(iris[1:5, ], caption = 'A caption')
```

**Obtain predicted values, residuals and anova of `lm()`**
```{r , message=FALSE}
# EARNINGS regress on schooling
library(dplyr)
eawe21 <- read.csv('./Data/EAWE21.csv', header=T)   # Import data
lm(EARNINGS ~ S, data = eawe21) %>%
  anova
```
(Please the website [https://rc2e.com/linearregressionandanova][rc2e] for more information on regression output.)

[rc2e]: https://rc2e.com/linearregressionandanova